Spatiotemporal Multi-Resolution Approximation of the Amari Type Neural Field Model

### 2012-01-31 Multiresolution_framework/ltx/manuscript.tex
added a section for stopping criterion

The algorithm has two steps: the E-step, which computes $\boldsymbol\Xi_0$ and $\boldsymbol\Xi_1$ based on the most recent parameter estimates using the RTS smoother, and the M-step, which updates the parameter estimates by computing the (analytic) maximum of $Q(\theta,\theta')$. The EM algorithm iterates between these two steps until the parameter estimates converge. The RTS  algorithm is given in Algorithm~\ref{alg:MRA-RTS} for completeness. 
\begin{algorithm}
\caption{Summary of the Rauch-Tung-Striebel smoother}
\label{alg:MRA-RTS}
 % \begin{tabbing}
 \begin{enumerate}
 \item Forward initialisation: 
 \begin{equation*}
 \hat{\mathbf x}_0, \mathbf P_0
 \end{equation*}
 \item Forward iteration: for $t\in\left\lbrace 0,\cdots,T\right\rbrace $, calculate the predicted state and the predicted covariance matrix
 \begin{align}
 \hat{\mathbf x}_{t+1}^{f-}&=\mathbf A \hat{\mathbf x}_{t}^{f} \nonumber \\
 \mathbf P_{t+1}^{f-}&=\mathbf A \mathbf P_{t}^{f}\mathbf A^{\top}+\boldsymbol\Sigma_e \nonumber 
 \end{align}
 \item Compute the filter gain, the filtered state and the filtered covariance matrix
 \begin{align}
 \mathcal K_{t+1}&=\mathbf P_{t +1}^{f-}\mathbf C ^\top(\mathbf C \mathbf P_{t +1}^{f-}\mathbf C ^\top+\boldsymbol \Sigma_{\varepsilon})^{-1} \nonumber\\
\hat{\mathbf x}_{t+1}^{f}&=\hat{\mathbf x}_{t+1}^{f-}+\mathcal K_{t+1}(\mathbf y_{t+1}-\mathbf C\hat{\mathbf x}_{t +1}^{f-})\nonumber \\
 \mathbf P_{t+1}^f&=(\mathbf I - \mathcal K_{t+1}\mathbf C)\mathbf P_{t +1}^{f-} \nonumber 
 \end{align}
 \item  Backward initialisation:
 \begin{equation*}
 \mathbf P_T^b= \mathbf P_T^f, \quad \hat{\mathbf x}^b_T= \hat{\mathbf x}^f_T
 \end{equation*}
 \item Backward iteration: for $t\in\left\lbrace T-1,\cdots,0\right\rbrace $ compute the smoother gain, the smoothed state and the smoothed covariance matrix
 \begin{align}
  \mathbf S_{t}&=\mathbf P_{t}^{f}\mathbf A^{\top}\left[ \mathbf P_{t +1}^{f-}\right]^{-1} \nonumber\\
 \hat{\mathbf x}_t^b&=\hat{\mathbf x}_t^f+\mathbf S_t(\hat{\mathbf x}_{t+1}^{b}-\hat{\mathbf x}_{t+1}^{f-})\nonumber \\
 \mathbf P_{t}^{b}&=\mathbf P_{t}^{f}+\mathbf S_t(\mathbf P_{t+1}^{b}-\mathbf P_{t+1}^{f-})\mathbf S_t^\top \nonumber
 \end{align}
 \item compute the smoothed cross covariance matrix\\
 initialisation:
 \begin{align}
  \mathbf M_T=(\mathbf I-\mathcal K_T\mathbf C)\mathbf A\mathbf P_{T-1}^b \nonumber
 \end{align}
 Backward iteration: for $t\in\left\lbrace T-1,\cdots,0\right\rbrace $
 \begin{align}
 \mathbf M_t= \mathbf P_t^{f}\mathbf S_{t-1}^{\top}+\mathbf S_{t}(\mathbf M_{t+1}-\mathbf A\mathbf P_t^{f} )\mathbf S_{t-1}^{\top}\nonumber
 \end{align}
 \end{enumerate}
 % \end{tabbing}
 \end{algorithm}
The stopping criterion is usually associated with either the change in the parameter estimates or the log-likelihood variation \cite{McLachlan1997}. The stopping rule adopted herein is
\begin{equation}
 \left(\parallel \mathbf{A} \parallel_{F}^{(i)}-\parallel \mathbf{A} \parallel_{F}^{(i-1)}\right)<\epsilon,
 \end{equation}
 where $\epsilon$ is a threshold value and $\parallel \mathbf{A} \parallel_{F}^{(i)}$ and $ \parallel \mathbf{A} \parallel_{F}^{(i-1)}$ are the Frobenius norms  of the successive estimates of $\mathbf{A} $ matrices which is defined by \cite{Meyer2000}
 \begin{equation}
  \parallel \mathbf{A} \parallel_{F}=\sqrt{\sum_{i,j=1}^{n_x}\mid a_{i,j} \mid^2}=\sqrt{\mathrm{tr} (\mathbf A^{\top}\mathbf A)}
 \end{equation}
%\cut{The second terms in  \eqref{eq:upsilon0}, \eqref{eq:upsilon1}, and \eqref{eq:Upsilon} can be computed as a once-off before the commencement of the EM iterations, which increases the speed of the M-step significantly compared to the method in \cite{Dewar2009}.}

>>>>

The algorithm has two steps: the E-step, which computes $\boldsymbol\Xi_0$ and $\boldsymbol\Xi_1$ based on the most recent parameter estimates using the RTS smoother, and the M-step, which updates the parameter estimates by computing the (analytic) maximum of $Q(\boldsymbol\theta,\boldsymbol\theta')$. The EM algorithm iterates between these two steps until the parameter estimates converge. The RTS  algorithm is given in Algorithm~\ref{alg:MRA-RTS} for completeness. 
\begin{algorithm}
\caption{Summary of the Rauch-Tung-Striebel smoother}
\label{alg:MRA-RTS}
 % \begin{tabbing}
 \begin{enumerate}
 \item Forward initialisation: 
 \begin{equation*}
 \hat{\mathbf x}_0, \mathbf P_0
 \end{equation*}
 \item Forward iteration: for $t\in\left\lbrace 0,\cdots,T\right\rbrace $, calculate the predicted state and the predicted covariance matrix
 \begin{align}
 \hat{\mathbf x}_{t+1}^{f-}&=\mathbf A \hat{\mathbf x}_{t}^{f} \nonumber \\
 \mathbf P_{t+1}^{f-}&=\mathbf A \mathbf P_{t}^{f}\mathbf A^{\top}+\boldsymbol\Sigma_e \nonumber 
 \end{align}
 \item Compute the filter gain, the filtered state and the filtered covariance matrix
 \begin{align}
 \mathcal K_{t+1}&=\mathbf P_{t +1}^{f-}\mathbf C ^\top(\mathbf C \mathbf P_{t +1}^{f-}\mathbf C ^\top+\boldsymbol \Sigma_{\varepsilon})^{-1} \nonumber\\
\hat{\mathbf x}_{t+1}^{f}&=\hat{\mathbf x}_{t+1}^{f-}+\mathcal K_{t+1}(\mathbf y_{t+1}-\mathbf C\hat{\mathbf x}_{t +1}^{f-})\nonumber \\
 \mathbf P_{t+1}^f&=(\mathbf I - \mathcal K_{t+1}\mathbf C)\mathbf P_{t +1}^{f-} \nonumber 
 \end{align}
 \item  Backward initialisation:
 \begin{equation*}
 \mathbf P_T^b= \mathbf P_T^f, \quad \hat{\mathbf x}^b_T= \hat{\mathbf x}^f_T
 \end{equation*}
 \item Backward iteration: for $t\in\left\lbrace T-1,\cdots,0\right\rbrace $ compute the smoother gain, the smoothed state and the smoothed covariance matrix
 \begin{align}
  \mathbf S_{t}&=\mathbf P_{t}^{f}\mathbf A^{\top}\left[ \mathbf P_{t +1}^{f-}\right]^{-1} \nonumber\\
 \hat{\mathbf x}_t^b&=\hat{\mathbf x}_t^f+\mathbf S_t(\hat{\mathbf x}_{t+1}^{b}-\hat{\mathbf x}_{t+1}^{f-})\nonumber \\
 \mathbf P_{t}^{b}&=\mathbf P_{t}^{f}+\mathbf S_t(\mathbf P_{t+1}^{b}-\mathbf P_{t+1}^{f-})\mathbf S_t^\top \nonumber
 \end{align}
 \item compute the smoothed cross covariance matrix\\
 initialisation:
 \begin{align}
  \mathbf M_T=(\mathbf I-\mathcal K_T\mathbf C)\mathbf A\mathbf P_{T-1}^b \nonumber
 \end{align}
 Backward iteration: for $t\in\left\lbrace T-1,\cdots,0\right\rbrace $
 \begin{align}
 \mathbf M_t= \mathbf P_t^{f}\mathbf S_{t-1}^{\top}+\mathbf S_{t}(\mathbf M_{t+1}-\mathbf A\mathbf P_t^{f} )\mathbf S_{t-1}^{\top}\nonumber
 \end{align}
 \end{enumerate}
 % \end{tabbing}
 \end{algorithm}
\subsection{Stopping rule}
The stopping criterion is usually associated with either the change in the parameter estimates or the log-likelihood variation \cite{McLachlan1997}. The stopping rule adopted herein is
\begin{equation}
 \left(\parallel \mathbf{A} \parallel_{F}^{(i)}-\parallel \mathbf{A} \parallel_{F}^{(i-1)}\right)<\epsilon,
 \end{equation}
 where $\epsilon$ is a threshold value and $\parallel \mathbf{A} \parallel_{F}^{(i)}$ and $ \parallel \mathbf{A} \parallel_{F}^{(i-1)}$ are the Frobenius norms  of the successive estimates of $\mathbf{A} $ matrices which is defined by \cite{Meyer2000}
 \begin{equation}
  \parallel \mathbf{A} \parallel_{F}=\sqrt{\sum_{i,j=1}^{n_x}\mid a_{i,j} \mid^2}=\sqrt{\mathrm{tr} (\mathbf A^{\top}\mathbf A)}
 \end{equation}
%\cut{The second terms in  \eqref{eq:upsilon0}, \eqref{eq:upsilon1}, and \eqref{eq:Upsilon} can be computed as a once-off before the commencement of the EM iterations, which increases the speed of the M-step significantly compared to the method in \cite{Dewar2009}.}
